Type Refinements for the Working Class

### 2015-12-20 paper.tex
add section on families of refinements

\newcommand\TyBase[1]{\mathtt{Base}_{#1}}
\newcommand\TyTop[1]{\top_{#1}}
\newcommand\TyBot[1]{\bot_{#1}}
\newcommand\TyMeet[1]{\cap_{#1}}
\newcommand\TyJoin[1]{\cup_{#1}}

@@

We have the bottom refinement $\GloballyRefines{\TyBot{\tau}}{\tau}$, which
simply contains no elements. Next, the top refinement
$\GloballyRefines{\TyTop{\tau}}{\tau}$ identifies all terms
$\Member{M,N}{\Exprs_\tau(\Upsilon)}$.  The join
$\Refines{\Upsilon}{\phi\TyJoin{\tau}\psi}{\tau}$ of two refinements
$\Refines{\Upsilon}{\phi}{\tau}$ and $\Refines{\Upsilon}{\psi}{\tau}$ is defined as
follows:

\begin{gather*}
  \infer{
    \RMemEq{\Upsilon}{M}{N}{\phi\TyJoin{\tau}\psi}
  }{
    \RMemEq{\Upsilon}{M}{N}{\phi}
  }\qquad
  \infer{
    \RMemEq{\Upsilon}{M}{N}{\phi\TyJoin{\tau}\psi}
  }{
    \RMemEq{\Upsilon}{M}{N}{\psi}
  }
\end{gather*}

Finally the meet $\Refines{\Upsilon}{\phi\TyMeet{\tau}\psi}{\tau}$ of two refinements
$\Refines{\Upsilon}{\phi}{\tau}$ and $\Refines{\Upsilon}{\psi}{\tau}$ is defined
dually:
\[
  \infer{
    \RMemEq{\Upsilon}{M}{N}{\phi\TyMeet{\tau}\psi}
  }{
    \RMemEq{\Upsilon}{M}{N}{\phi} &
    \RMemEq{\Upsilon}{M}{N}{\psi}
  }
\]

@@

>>>>

\newcommand\TyBase[1]{\mathtt{Base}_{#1}}
\newcommand\TyTop[1]{\top_{#1}}
\newcommand\TyBot[1]{\bot_{#1}}

@@

We have the bottom refinement $\GloballyRefines{\TyBot{\tau}}{\tau}$, which
simply contains no elements. Next, the top refinement
$\GloballyRefines{\TyTop{\tau}}{\tau}$ identifies all terms
$\Member{M,N}{\Exprs_\tau(\Upsilon)}$.  The join (union)
$\Refines{\Upsilon}{\phi\cup\psi}{\tau}$ of two refinements
$\Refines{\Upsilon}{\phi}{\tau}$ and $\Refines{\Upsilon}{\psi}{\tau}$ is defined as
follows:

\begin{gather*}
  \infer{
    \RMemEq{\Upsilon}{M}{N}{\phi\cup\psi}
  }{
    \RMemEq{\Upsilon}{M}{N}{\phi}
  }\qquad
  \infer{
    \RMemEq{\Upsilon}{M}{N}{\phi\cup\psi}
  }{
    \RMemEq{\Upsilon}{M}{N}{\psi}
  }
\end{gather*}

Finally the meet (intersection) $\Refines{\Upsilon}{\phi\cap\psi}{\tau}$ of two refinements
$\Refines{\Upsilon}{\phi}{\tau}$ and $\Refines{\Upsilon}{\psi}{\tau}$ is defined
dually:
\[
  \infer{
    \RMemEq{\Upsilon}{M}{N}{\phi\cap\psi}
  }{
    \RMemEq{\Upsilon}{M}{N}{\phi} &
    \RMemEq{\Upsilon}{M}{N}{\psi}
  }
\]

@@

\subsection{Families of Refinements}

Given a family of refinements (i.e. a functional refinement)
$\FRefines{\Upsilon}{x:\phi}{\psi}{\tau}$ such that
$\Refines{\Upsilon}{\phi}{\sigma}$, including the family intersection, the
dependent intersection, the family union, and the set comprehension.

\paragraph{Family intersection}
The intersection $\Refines{\Upsilon}{\bigcap_{(x:\phi)}\psi}{\tau}$ of a family
of refinements is defined by the following rule, for any
\begin{tikzcd}[cramped, ampersand replacement = \&]
  \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] \&\IMode{\Upsilon'}
\end{tikzcd}:
\[
  \infer={
    \RMemEq{\Upsilon'}{M}{N}{
      \bigcap_{(x:\phi)}\psi
    }
  }{
    \FRMemEq{\Upsilon'}{x:\phi}{M}{N}{\Subst{M}{x}{\psi}}
  }
\]

The intersection of a family of refinements is a kind of uniform universal
quantification, in the sense that the terms themselves bear no trace of the
quantification, but are guaranteed in the metalogic to be related in all fibres
of $\psi$.

In practice, the family intersection is perhaps the most-used type constructor
in Nuprl developments, as it facilitates the simultaneous specification and
implementation of programs in type theory, without causing details of the
program's verification to leak into the extracted algorithm. Additionally, the
family intersection can be used to implement coinductive types.

\paragraph{Dependent intersection}
The dependent intersection $\Refines{\Upsilon}{(x:\phi)\cap\psi}{\sigma}$
contains the members $M$ of $\phi$ which are also members of
$\Subst{M}{x}{\psi}$:
\[
  \infer={
    \RMemEq{\Upsilon'}{M}{N}{
      (x:\phi)\cap\psi
    }
  }{
    \RMemEq{\Upsilon'}{M}{N}{\phi} &
    \RMemEq{\Upsilon'}{M}{N}{
      \Subst{M}{x}{\psi}
    }
  }
\]

The dependent intersection, first introduced by Kopylov in \cite{kopylov:2000},
are a generalization of the binary intersection to the dependent case, not to
be confused with the intersection of a family of sets. The dependent
intersection has many applications, including the compositional representation
of \emph{dependent records} or modules in Type Theory.

\paragraph{Family union}
The union $\Refines{\Upsilon}{\bigcup_{(x:\phi)}\psi}{\tau}$ of a family of
refinements is defined as by the following rule:
\[
  \infer{
    \RMemEq{\Upsilon'}{M}{N}{
      \bigcup_{(x:\phi)}\psi
    }
  }{
    \RMemEq{\Upsilon'}{L_0}{L_1}{\phi} &
    \RMemEq{\Upsilon'}{M}{N}{
      \Subst{L_0}{x}{\psi}
    }
  }
\]

The union of a family of refinements can be used to implement inductive types,
among other things.

\paragraph{Set comprehension}

We can also use a family of refinements to define something analogous to set
comprehension, $\Refines{\Upsilon'}{\MkSet{x:\phi\mid\psi}}{\sigma}$, defined
as follows:
\[
  \infer{
    \RMemEq{\Upsilon'}{M}{N}{
      \MkSet{x:\phi\mid\psi}
    }
  }{
    \RMemEq{\Upsilon'}{M}{N}{\phi} &
    \RMemEq{\Upsilon'}{L_0}{L_1}{
      \Subst{M}{x}{\psi}
    }
  }
\]

The set comprehension $\MkSet{x:\phi\mid\psi}$ contains just those elements of
$\phi$ such that their fibres of $\psi$ are inhabited; this can be seen as an
``implicit'' version of existential quantification, and can be used to
simultaneously specify and implement program without the evidence for $\psi$
intruding into the extracted algorithm.
